---
title: "Bayesian Structural Time Series"
---

This post is inspired by these two posts, [Fitting Bayesian structural time series with the bsts R package](http://www.unofficialgoogledatascience.com/2017/07/fitting-bayesian-structural-time-series.html) and a subsequent post on the Stan help page [here](https://discourse.mc-stan.org/t/bayesian-structural-time-series-modeling/2256).

```{r}
library('rstan')
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
library(dplyr)
```


## Data Generating Process

Regression State
$$y_t = \mu_t + \beta^Tx_t + \epsilon_t$$

Trend State

$$\mu_{t+1}=\mu_t+\delta_t+\eta_{0t}$$


```{r}
set.seed(336)
x1 <- rnorm(250,1,0.1)
x2 <- rnorm(250, 0, 1)
x3 <- rnorm(250, 1, .5)

preds <- data.frame(x1 = x1, x2 = x2, x3 = x3)
v <- rep(-0.0001,250)
u <- rep(0.0,250)
s_slope <- rnorm(250,0,0.005)
s_level <- rnorm(250,0,0.005)
for (i in 2:250){
  u[i] <- u[i-1] + v[i-1] + s_level[i-1]
  v[i] <- v[i-1] + s_slope[i-1]
}

s_obs <- rnorm(250,0,1)
x <- runif(250,0,1)
y <- u + 5*x1 + 1*x2 + -.75*x3+ s_obs
n <- 250
x <- as.matrix(preds)
k <- ncol(preds)
ytrain <- y[1:n]
xtrain <- x[1:n,]

plot(y)

```

## Stan

Develop the Stan model

```{r stan-bsts, comment=""}
writeLines(readLines("stan_bsts.stan"))
```


```{r}
stan_dat = list( t = length(ytrain), 
                 y = ytrain, 
                 x = xtrain,
                 K = k)
```

Compile the Stan model.

```{r}
model_bsts <- stan_model("stan_bsts.stan")
```

The run the model with our data.

```{r}
#Load stan model file, and fit to data
fit <- sampling(model_bsts , stan_dat,
                iter=1000, 
                chains=3, 
                control = list(max_treedepth = 15,
                               adapt_delta = .95), refresh =0)

```

## Model Checking

And as always special thanks to [Michael Betancourt](https://betanalpha.github.io/) for these amazing tools for diagnostics. 

```{r check-tools}
util <- new.env()
source('stan_utilities.R', local=util)
```

```{r checks, eval=FALSE}
util$check_all_diagnostics(fit)
```

### Do pairs plot of different parameters

```{r}
pairs(fit,pars = c("s_slope","s_level","s_obs","beta"))

```

```{r}
traceplot(fit, pars = "beta")
```


## Inference

```{r}
print(fit, pars = "beta")
```

## Add Seasonality?

### Data Generating Process

Regression State
$$y_t = \mu_t + \tau_t + \beta^Tx_t + \epsilon_t$$

Trend State

$$\mu_{t+1}=\mu_t+\delta_t+\eta_{0t}$$

Seasonal Pattern

$$\tau_{t+1}=-\Sigma_{s=1}^{s-1}\tau_t+\eta_{2t}$$

## Generate Fake Data

```{r}
set.seed(336)
x1 <- rnorm(250,1,0.1)
x2 <- rnorm(250, 1, .1)
x3 <- rnorm(250, 1, .5)

preds <- data.frame(x1 = x1, x2 = x2, x3 = x3)
v <- rep(-0.0001,250)
u <- rep(0.0,250)
s_slope <- rnorm(250,0,0.005)
s_level <- rnorm(250,0,0.005)

for (i in 2:250){
  u[i] <- u[i-1] + v[i-1] + s_level[i-1]
  v[i] <- v[i-1] + s_slope[i-1]
}

# Seasonal Component with freq = 12
t <- seq(0,1, by = 1/250)

tau <-vector()

tau<- 1*sin(12*pi*t*2)

plot(tau)
s_obs <- rnorm(250,0,1)
x <- runif(250,0,1)
y <- u + 5*x1 + 1*x2 + -.75*x3+ s_obs + tau[-251]
n <- 250
x <- as.matrix(preds)
k <- ncol(preds)
ytrain <- y[1:n]
xtrain <- x[1:n,]

plot(y)
```

## Stan Model

```{r stan-bsts, comment=""}
writeLines(readLines("stan_bsts_seasonal.stan"))
```

Now let's compule the model:

```{r}
model_season <- stan_model("stan_bsts_seasonal.stan")
```

Prepare the data  for Stan.

```{r}
stan_dat_seasonal = list( t = length(ytrain), 
                 y = ytrain, 
                 x = xtrain,
                 K = k,
                 frequency = 12)
```

Then fit the model.

```{r}
fit <-sampling(model_season, stan_dat_seasonal, 
               iter = 1000, cores = 2, chain = 2,
               control = list(max_treedepth = 15, adapt_delta = .99), 
               refresh = 0)
```

```{r checks, eval=FALSE}
util$check_all_diagnostics(fit)
```


```{r}
print(fit, "beta")
```

```{r}
pairs(fit, pars = c("beta", "s_obs", "s_season"))
```

