---
title: "Confirmatory Factor Analysis in Stan"
---

```{r setup, include=FALSE}
library(tidyverse, quietly = T)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```


The motivation for this post is to learn how to perform [onfirmatory factor analysis](https://en.wikipedia.org/wiki/Confirmatory_factor_analysis)(CFA) in Stan (and thus in a Bayesian way). These measures are typically noisey so Bayes, in my opinion, lends itself well to these kind of constructs. Additionaly, the gold standard for CFA in R is `lavaan`. This is nice because it provides a decent benchmark for the output of the Bayesian version. This post is largely inspired by the work done by Mike Lawrence on this topic [here](https://gist.github.com/mike-lawrence/dd2435f290a567bd1fd03370ee669688).

## Data

The data that will be used will be that from the `HolzingerSwineford1939` dataset that is part of the `lavaan` package.

```{r}
library(lavaan)

HolzingerSwineford1939
```


## Hypothesis Testing

CFA differs from exploratory factor analysis (EFA) in the fact that one has a hypothesis for how the _factors_ of a latent trait form. So the first step is to develop the hypothesised model:

```{r sem-plot, echo=FALSE}
cfa_model <- ' visual  =~ x1 + x2 + x3 
              textual =~ x4 + x5 + x6
speed   =~ x7 + x8 + x9 '

cfa_fit <- cfa(cfa_model, HolzingerSwineford1939)
semPlot::semPaths(cfa_fit)
```

## Implmentation

### Maximum Likelihood

Just to give the reference, first we will conduct the analysis in `lavaan`

```{r mle-cfa-fit}
cfa_model <- ' visual  =~ x1 + x2 + x3 
              textual =~ x4 + x5 + x6
speed   =~ x7 + x8 + x9 '

cfa_fit <- cfa(cfa_model, HolzingerSwineford1939)
```

And we can examine the outputs:

```{r mle-cfa-sum}
summary(cfa_fit, fit.measures=TRUE)
```

## Bayesian Version

Now first to write the Stan code:

```{r stan-1}
library(rstan)
rstan_options(auto_write = TRUE)
writeLines(readLines("stan_cfa.stan"))
```

Then compile the model:

```{r compile}
stan_cfa_model <- stan_model("stan_cfa.stan")
```


### Prep the Data

Of course with Stan we have to format our data. One note here is that we have to specify the parameters for our latent traits as well.

```{r prep-data}
HolzingerSwineford1939 %>%
	dplyr::select(
		x1,x2,x3,x4,x5,x6,x7,x8,x9
	) -> dat

stan_dat <- list(
	# n_y: number of outcomes
	n_y = ncol(dat)
	# n_subj: number of subjects
	, n_subj = nrow(dat)
	# y: matrix of outcomes
	, y = as.matrix(dat)
	# n_fac: number of latent factors
	, n_fac = 3
	# y_fac: list of which factor is associated with each outcome
	, y_fac = c(1,1,1,2,2,2,3,3,3)
)
```

Then run the model

```{r run-stan}
stan_fit <- sampling(stan_cfa_model, stan_dat, 
                     cores = 3, chains = 3, iter = 4000,
                     refresh = 0)
```

Let's examine the convergence to make sure that enough iterations have been run.

```{r checks}
traceplot(stan_fit)
```


Not too bad. Now we can review our parameters

```{r posterior}
print(stan_fit, pars = c("betas", "fac_cor", "y_means", "y_noise"))
```

