---
title: "Linear Regression in Stan"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## My Mission

Write a bunch of Stan Code to practice building models in Stan. While `brms` is fantastic, I feel like I need to put in the work to learn to build the Stan code myself. I am going to try and work through one of the exercises each day from the [handbook](https://mc-stan.org/docs/2_18/stan-users-guide/index.html) and other websites dealing with interesting models.

## Linear Regression

```{r}
library(rstan)
```

Create Some Basic Data:

```{r}
x <- rnorm(40, 10, 5)
noise <- rnorm(40,0,1)
y <- x*.5 + noise

data <- list( x= x, y = y, N = length(x))
```

The below Stan code models the familiar $y = \beta*x + \alpha + \epsilon$

Or more formally:

$$y_n \sim N(\alpha + \beta X_n,\sigma)$$

```
data {
  int<lower=0> N;
  vector[N] x;
  vector[N] y;
}
parameters {
  real alpha;
  real beta;
  real<lower=0> sigma;
}
model {
  y ~ normal(alpha + beta * x, sigma);
}
```

So now we need to compile the Stan code. This takes a little while....

```{r}
linear_regression <- stan_model("stan_linear_regression.stan")
```

One that code has been compiled then we can actually fit the model. This is a simple model and it converges quickly (which it should).

```{r}
fit1 <- sampling(linear_regression, data = data, chains = 2, iter = 1000, refresh = 0)
```

Now we can look at the model outputs with the `summary` which prints a lot of information or just look at the parameters one by one. I'm interested in if it could detect the true slope, in this case 0.5:

```{r}
print(fit1, pars = "beta", probs = c(0.025, 0.5, 0.975))
```

And it did a reasonable job! Yahoo it worked.

As with any good Bayesian analysis it is important to perform some posterior checks to ensure that the model sufficiently converged. One check is the $\hat{R}$ which is a measure of the mixing on the chaines with a target of one (which is achieved here).

Additionally, we can look at the trace plots to make sure everything converged and thise look good too.

```{r}
traceplot(fit1)
```

And finally a pairs plot which shows that sigma, alpha and beta all look reasonable centered with no strange patterns.

```{r}
pairs(fit1)
```

So, yahoo! first real model in Stan!
