---
title: "Autoregressive Modeling in Stan"
---

This post looks at modeling autoregressive models. Autoregression is typically important when dealing with time series analysis. The general idea is that the past value of a repeatedly measured item will be indicative of its future value. A good example is a stock. Today's stock price could be modelled as a function of yesterday's stock price. While this isn't entirely true, it does approximate a decent model.


## Data Generating Process

In this case I am going to look at an AR(3) model. Thus the model form would take on the following data generating process:

$$y_n \sim N(\alpha + \beta_1y_{n-1} + \beta_2y_{n-2} + \beta_3y_{n-3}, \sigma)$$

And we can used a built in feature of R to help us simulate these data. Note that I have not specificed a moving average. 

```{r}
fake_data <-arima.sim(n = 200, model = list(ar = c(.2, .5, .05)))
```


```{r}
ts.plot(fake_data, main = "Time Series Plot of Our Fake Data")
```


## Stan

### The Model

```{r comment=""}
writeLines(readLines("stan_ar_models.stan"))
```

### The Modeling Process

```{r}
library(rstan)

rstan_options(auto_write = TRUE)
model <- stan_model("stan_ar_models.stan")

```

### Format Data

Now we need to format our data for our Stan program:

```{r}

stan_dat <- list(
  N = length(fake_data),
  K = 3,
  y = as.vector(fake_data)
)

```

### Fit the Model

Let's run our model with the usual conditions.

```{r cache=TRUE}
fit <- sampling(model, stan_dat,
                cores = 2, iter = 1000,
                refresh = 0, chains = 2)
```

## Model Checking

And as always special thanks to [Michael Betancourt](https://betanalpha.github.io/) for these amazing tools for diagnostics. 

```{r check-tools}
util <- new.env()
source('stan_utilities.R', local=util)
```

```{r checks, eval=FALSE}
util$check_all_diagnostics(fit)
```

## Inference

```{r}
print(fit, pars = "beta")
```

