---
title: Lectures
bibliography: library.bib
biblio-style: apalike
---

```{r, child="_setup.Rmd"}
```

## Introduction to R

### tldr:
>An overview of R, RStudio and github. The focus is on orienting users to R via R Studio and getting comfortable within the IDE.


```{r echo=FALSE, results='asis'}
make_doc_links("introduction")
```

## JHU Advanced Datascience

<https://jhu-advdatasci.github.io/2018/>

## Docker with R

<https://colinfay.me/docker-r-reproducibility/>

## R MArkdown for Scientists

<https://rmd4sci.njtierney.com/>

## R Metapackage

<http://pkgverse.mikewk.com/>

## .gitignore template

<https://github.com/github/gitignore/blob/master/R.gitignore>

## Rmarkdown website

<https://stanford.edu/~vbauer/teaching/websites.html#faking-website-organization>

## Shareable Link to Data R

<https://gist.github.com/alexpghayes/9118cda66375e593343fe28c8d13fdb5>

## Publishing with GH and Travis

<https://www.r-bloggers.com/publishing-from-blogdown-to-github-with-travis/>

## Adding Firebase Authentication to Shiny

<https://www.tychobra.com/posts/2019-01-03-firebasse-auth-wtih-shiny/>